A Way Forward in Parallelising Dynamic Languages

### 2014-06-18 talk/icooolps2014/position-paper.tex
some more adjustments

@@

\begin{description}
\item[Performance:] How much does the approach impact performance on a single
  and how much on multiple threads? Can it make use of parallelism?
\item[Existing applications:] How big are the changes required to
  integrate with and parallelise existing applications?
\item[Better synchronisation:] Does the approach enable better,
  paralleliseable synchronisation mechanisms for applications
  (e.g.\ atomic blocks)?  Many synchronisation mechanisms can be built on
  top of all solutions (e.g.\  message passing). We look for mechanisms
  that are directly enabled by the contending approaches.
\item[Implementation:] How difficult is it to implement the approach
  in the interpreter?
\item[External libraries:] Does the approach allow for easy
  integration of external libraries?
\end{description}

@@

However, STM compared to HTM does not suffer from the same restricting
limitations. Transactions can in principle be arbitrarily long.  This makes it
possible to actually expose transactions to the application in the
form of atomic blocks. This is the only approach that enables a better
synchronisation mechanism than locks for applications \emph{and} still
parallelises when using it. We think this is a very important point
because it not only gives dynamic languages the ability to parallelise
(already commonplace in most other languages), but also pushes
parallel programming forward. Together with sequential consistency it
provides a lot of simplification for parallel applications.

@@

Following the above argumentation for each approach, we assembled a
general overview in Table \ref{tab:comparison}. The points were assigned
according to the criteria described in \ref{sec:pot_solutions}. Since
the criteria are defined intuitively, there are no formal justifications
for the number of points. The reader is thus advised to take the result
with a grain of salt.

@@

Allowing for parallel execution just means that dynamic languages
catch up to all other languages that already provide real
parallelism. This is why we think that only the STM approach is a
viable solution in the long-term. It unifies both, the simple memory
model (sequential consistency) and the synchronisation of memory accesses
using composable atomic blocks. It is not \emph{only} a simple GIL
replacement.

Unfortunately, STM has a big performance problem. Particularly, for
our use case there is not much static information available since we
are executing a program only known at runtime. Additionally, replacing
the GIL means running every part of the application in transactions,
so there is not much code that can run outside and that can be
optimised better. The performance of the TM system is vital.

One way to get more performance is to develop STM systems that make
better use of low-level features in existing OS kernels.  We are
currently working on an STM system that makes use of several such
features like virtual memory and memory segmentation.  We further
tailor the system to the discussed use case, which gives us an
advantage over other STM systems that are more general. With this
approach, initial results suggest that we can keep the overhead of STM
well below 50\%. A hybrid TM system, which also uses HTM to accelerate
certain tasks, looks like a very promising direction of research
too.

We think that further work to reduce the overhead of STM is
very worthwhile. In fact, considering some analogies that have been
drawn between garbage collection and transactional memory~\cite{dan07},
we think that it is worthwhile to focus the STM research more
specifically onto the context shown in this paper --- for use in
implementations of high-level languages, rather than as a tool
directly used by the programmer.

>>>>

% Keine "Schusterjungen"
\clubpenalty = 10000
% Keine "Hurenkinder"
\widowpenalty = 10000 \displaywidowpenalty = 10000

@@

\begin{description}
\item[Performance:] How much does the approach impact performance on a single
  and how much on multiple threads? Can it make use of parallelism?
\item[Existing applications:] How big are the changes required to
  integrate with and parallelise existing applications?
\item[Better synchronisation:] Does the approach enable better,
  parallelisable synchronisation mechanisms for applications
  (e.g.\ atomic blocks)?  Many synchronisation mechanisms can be built on
  top of all solutions (e.g.\  message passing, monitors). We look for mechanisms
  that are directly enabled by the contending approaches.
\item[Implementation:] How difficult is it to implement the approach
  in the interpreter?
\item[External libraries:] Does the approach allow for easy
  integration of external libraries?
\end{description}

@@

However, STM compared to HTM does not suffer from the same restricting
limitations. Transactions can in principle be arbitrarily long.  This makes it
possible to expose transactions to the application in the
form of atomic blocks -- thereby attacking the issues
of parallelisation and synchronisation in a unified way. While many
synchronisation mechanisms can be bolted on top of any GIL replacement,
this is the only approach that directly enables a better, parallelising
synchronisation mechanism than locks.  We think this is a very
important point because it not only gives dynamic languages the
ability to parallelise (already commonplace in most other languages),
but also pushes parallel programming forward in a way that other
approaches cannot. Together with sequential consistency, it provides an
environment for parallel applications that has much less surprises than
e.g.\ Java or C\#.

@@

Following the above argumentation for each approach, we assembled a
general overview in Table \ref{tab:comparison}. The points were assigned
according to the criteria described in \ref{sec:pot_solutions}. Since
the criteria are defined intuitively, there are no formal justifications
for the number of points. The reader is thus advised to take the result
with a grain of salt and form their own opinion.

@@

Allowing for parallel execution just means that dynamic languages
catch up to all other languages that already provide real
parallelism. This is why we think that only the STM approach is a
viable solution in the long-term. It unifies both, the simple memory
model (sequential consistency) and the synchronisation of memory accesses
using composable atomic blocks. It is not \emph{just} a simple GIL
replacement.

Unfortunately, STM has a big performance problem, which currently
makes it lose this comparison. Particularly, for
our use case there is not much static information available since we
are executing a program only known at runtime. Additionally, replacing
the GIL means running every part of the application in transactions,
so there is not much code that can run outside and that can be
optimised better. The performance of the TM system is vital.

One way to get more performance is to develop STM systems that make
better use of low-level features in existing OS kernels.  We are
currently working on an STM system that makes use of several such
features like virtual memory and memory segmentation.  We further
tailor the system to the discussed use case, which gives us an
advantage over other STM systems that try to be more general or
simply focus on other use cases. With this
approach, initial results suggest that we can keep the overhead of STM
well below 50\%. A hybrid TM system, which also uses HTM to accelerate
certain tasks, looks like a very promising direction of research
too.

We think that further work to reduce the overhead of STM is
very worthwhile. In fact, considering some analogies that have been
drawn between garbage collection and transactional memory~\cite{dan07},
we think that it is worthwhile to focus the STM research more
specifically onto the context shown in this paper -- for use in
implementations of high-level languages, rather than as a tool
directly used by the programmer.
